---
title: "Session 3 - Manipulating Data with dplyr and tidyr"
author: "Alex Mounsey"
date: "11/16/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
```

## Exercise 1: Bank Data

The file `bank.xlsx` contains information regarding the customers of a bank, 
such as their town of residence, occupation, gender, and initial salary _(i.e. 
when they opened their account)_ and current salary in GBP _(£)_.

**Import `bank.xlsx` into R, naming the resulting R object `bank_data`. Take a 
look at the first few rows of `bank_data`:**
```{r}
bank_data <- read_excel('../data/bank.xlsx')
head(bank_data)
```

Note that the result is a 'tibble' _(or tidy table)_, which is a clever type of 
data frame.

**Create a new variable, `diff_sal`, as the difference between the current and 
initial salary:**
```{r}
bank_data <- bank_data %>%
  mutate(diff_sal = curr_sal - init_sal)

head(bank_data, 1)
```

**Considering only males, calculate the mean of `diff_sal` for each job 
category and show how many males are in each job category:**
```{r}
filter(bank_data, gender == 'M') %>%
  group_by(job) %>%
  summarise(count = n(), mean_diff_sal = mean(diff_sal))
```

**Create a new categorical variable, `curr_sal_new`, which takes the value 'high' 
when `curr_sal` is greater than £20,000 and the value 'low' otherwise:**
```{r}
bank_data <- bank_data %>%
  mutate(curr_sal_new = ifelse(curr_sal > 20000, 'high', 'low'))

bank_data[c(1, 6),]
```

**Selecting only the variables `town`, `gender`, `init_sal`, and `curr_sal`, 
reshape the data frame to obtain one value indicating the type of salary 
_(`sal_type`)_ and one variable indicating the value of the salary 
_(`sal_value`)_. Finally, order the data frame by `sal_value` in ascending order:**
```{r}
bank_data <- bank_data %>%
  gather('sal_type', 'sal_value', 4:5) %>%
  select(town, gender, sal_type, sal_value) %>%
  arrange(sal_value)

head(bank_data)
```

## Exercise 1.2: Other Graphical Displays of the Data

**Produce and comment a set of histograms depicting `sal_value`, faceted by 
`gender` and `sal_type`:**
```{r, message=FALSE}
# Rename variables for better readability
bank_data <- bank_data %>%
  mutate(sal_type = factor(sal_type,
                           levels = c('init_sal', 'curr_sal'),
                           labels = c('Initial_Salary', 'Current_Salary')))

# Plot salary by gender and salary type
ggplot(bank_data, aes(x = sal_value)) +
  theme_light() + geom_histogram() +
  facet_grid(gender ~ sal_type) +
  labs(x = "Salary (£)", y = "Count",
       title = "Salary Value",
       subtitle = "Split by gender and salary type")
```

### Comments:

Salary for both males and females have become more distributed _(males more so)_. 
At a glance, it appears that there are more males in the sample than females. Males 
appear to have a higher average salary than females.

**Produce and comment on a set of boxplots depicting `sal_type` against 
`sal_value`, faceted by `gender`:**
```{r}
ggplot(bank_data, aes(x = sal_type, y = sal_value)) +
  theme_light() + geom_boxplot() +
  facet_grid(. ~ gender) +
  labs(x = "Type of Salary", y = "Salary (£)",
       title = "Type of Salary vs. Salary Value",
       subtitle = "Split by gender")
```

## Exercise 2: 'Sailing & Dreams' Customer Satisfaction

A UK maritime transport company runs a transfer service between the UK and France. 
It has recently launched a new line of ferries, called 'Sailing & Dreams', 
offering many new services to passengers. In order to evaluate satisfaction 
towards this new line, the company has collected data from its passengers by 
means of an interview process.

The data is provided in three separate `.csv` files, which include the following 
information:

* **`PersonalInfo.csv`**, containing the variables:
  * _ID_: The passenger's ID
  * _Gender_: The passenger's gender (1: male, 2: female)
  * _Age_: The passenger's age (in years)
  * _Job_: The passenger's type of occupation
* **`PassengersInfo.csv`**, containing the variables:
  * _ID_: The passenger's ID
  * _FirstTime_: Whether the passenger is traveling with the company for the first 
  time (0: no, 1: yes)
  * _WorkHoliday_: Whether the passenger is traveling on holiday, or for work 
  (0: holiday, 1: work)
  * _Price_: The price of the trip in GBP (£)
  * _Propensity_: Whether the passenger intends to travel again on the 'Sailing 
  & Dreams' line (1: yes, 2: no)
* **Questionnaire.csv**, containing the passenger's ID _(ID)_ and the answers 
to the 14 questions of a questionnaire about their satisfaction with the service; 
answers to each of the questions are values between `0` and `5`, where:
  * `0`: Poor
  * `1`: Unsatisfactory
  * `2`: Acceptable
  * `3`: Satisfactory
  * `4`: Good
  * `5`: Excellent

**Use the information in the file `PassengersInfo.csv` to calculate interesting 
summary statistics about the passengers, split down into sensible groups such as 
the mean, variance, standard deviation, minimum and maximum price paid, travel 
times, and travel purpose:**
```{r, message=FALSE}
passenger_data <- read_csv('../data/PassengersInfo.csv')

passenger_data %>%
  group_by(FirstTime, WorkHoliday) %>%
  summarise(avg_price = mean(Price), var_price = var(Price), sd_price = sd(Price), 
            min_price = min(Price), max_price = max(Price))
```

**Create a single data frame, merging the three files: `PersonalInfo.csv`, 
`PassengersInfo.csv`, and `Questionnaire.csv` on the primary key `ID`, retaining 
only the rows present in all data frames:**
```{r, message=FALSE}
personal_data <- read_csv('../data/PersonalInfo.csv')
question_data <- read_csv('../data/Questionnaire.csv')

combined_data <- inner_join(personal_data, passenger_data, by = 'ID')
combined_data <- inner_join(combined_data, question_data, by = 'ID')

combined_data
```

**Create a new variable, `Job_2`, which combines the categories of `Job` into 
_`student`, `professional`, `not_working`, and `retired`_, using `ifelse()`. 
Then, tabulate the resulting `Job_2` variable:**
```{r}
combined_data <- combined_data %>%
  mutate(Job_2 = 
           ifelse(Job == 'howsewife', 'not_working',
           ifelse(Job == 'unemployed', 'not_working',
           ifelse(Job == 'retired', 'retired',
           ifelse(Job == 'student', 'student',
                  'professional')))))

table(combined_data$Job_2)
```

**Transform the variables `Propensity`, `FirstTime`, `WorkHoliday`, `Job_2`, and 
`Gender` into factors:**
```{r}
combined_data <- combined_data %>%
  mutate(Job_2_f = factor(Job_2)) %>%
  mutate(WorkHoliday_f = factor(WorkHoliday, labels = c('Holiday', 'Work'))) %>%
  mutate(Propensity_f = factor(Propensity, labels = c('Yes', 'No'))) %>%
  mutate(FirstTime_f = factor(FirstTime, labels = c('No, Yes'))) %>%
  mutate(Gender_f = factor(Gender, labels = c('Male', 'Female')))
```

**Produce a histogram of `Price`, coloured according to `Propensity`, and faceted 
by `FirstTime` and `WorkHoliday`:**
```{r, message=FALSE}
ggplot(combined_data, aes(x = Price, fill = Propensity_f)) +
  theme_light() + geom_histogram() +
  facet_grid(WorkHoliday_f ~ FirstTime_f) +
  labs(x = "Trip Price (£)", y = "Count", fill = "Propensity",
       title = "Histograms of Trip Price",
       subtitle = "Faceted by WorkHoliday and FirstTime, filled according to Propensity")
```

**Produce a boxplot of `Age`, stratified by `Job_2`, and faceted by `Gender`:**
```{r}
ggplot(combined_data, aes(x = Job_2_f, y = Age, fill = Job_2_f)) +
  theme_light() + geom_boxplot(varwidth = T) +
  facet_grid(. ~ Gender_f) +
  labs(x = "Occupation", y = "Age (years)", fill = "Occupation",
       title = "Age Stratified by Occupation and Gender") +
  theme(axis.text.x = element_text(size = 6.5, angle = 75, vjust = 0.5))
```

**Create a global satisfaction indicator, names `Score`, being the sum of the 
scores of all 14 questions:**
```{r}
combined_data <- combined_data %>%
  mutate(Score = PortCleanliness + PortComfort + PortStaff + Security + Accessibility +
           Disabled + Cost + SeatAvailability + JourneyTime + CleanlinessOnBoard + 
           ComfortOnBoard + StaffOnBoard + ServiceOnBoard + FoodOnBoard)
```

**Create a boxplot of `Score`, stratified by `Gender`, and faceted by `WorkHoliday` 
and `FirstTime`:**
```{r}
ggplot(combined_data, aes(x = Gender_f, y = Score, fill = Gender_f)) +
  theme_light() + geom_boxplot() +
  facet_grid(WorkHoliday_f ~ FirstTime_f) +
  labs(x = "Gender", y = "Score",
       title = "Score Stratified by Gender, WorkHoliday, and FirstTime")
```
