---
title: "Session 2 - Reading Data Manipulation"
author: "Alex Mounsey"
date: "30/09/20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(ggplot2)
```

## The Student Questionnaire Dataset

The file `MATH513_Questionnaire_Data.csv` contains information regarding a group 
of University of Plymouth students. A transcription error has been introduced 
for illustration purposes. Your task is to read in the data, perform data 
manipulation, and produce appropriate plots of specific variables.

**Read in the data from the file `MATH513_Questionnaire_Data.csv` using the 
function `read_csv()` from the `readr` package, and take a look at the data:**
```{r, message=FALSE}
q_data <- read_csv('../data/MATH513_Questionnaire_Data.csv')
head(q_data)
```

Suppose you are helping a team of health scientists that are studying the eating 
and drinking habits of the group of students. **Show the height, age, sex, and 
sports habits of the students who eat meat, drink coffee, and like beer:**
```{r}
filtered_q_data <- q_data %>%
  filter(EatMeat == 'Yes', DrinkCoffee == 'Yes', LikeBeer == 'Yes')

select(filtered_q_data, Height, Age, Sex, Sports)
```

The company which manages student accommodation is interested in analysing 
feedback about the quality of its services. Suppose now that you are helping the 
accommodation company. **Show interesting summary statistics about the 
interviewed students, such as the average and minimum `Sleep` time, and the median 
and maximum `Rent`. Split the results by students who are happy/not happy with 
their accommodation and those who do/don't feel safe. Additionally, show the 
number of students in each category. Comment on these results:**
```{r, message=FALSE}
q_data_by_happiness <- q_data %>%
  group_by(Happy_accommodation, Safe)

summarise(q_data_by_happiness,
          avg_sleep = mean(Sleep), min_sleep = min(Sleep),
          med_rent = median(Rent), max_rent = max(Rent),
          n_students = n())
```

## Dealing with Anomalous Points: Two Alternatives using `dplyr`

**Consider the entire questionnaire dataset. How many females and how many males 
are there?**
```{r}
count(q_data, vars = Sex)
```

**Produce boxplots of `Age`, stratified by `Sex`, using `geom_boxplot()`:**
```{r}
ggplot(q_data, aes(x = Sex, y = Age, fill = Sex)) +
  theme_light() + geom_boxplot(varwidth = 1) +
  labs(x = "Sex", y = "Age (years)",
       title = "Distribution of Age by Sex") +
  theme(legend.position = 'none')
```

There is clearly a problem with the data in that no student is under 15 years old. 
**Use `dplyr` to work out the minimum `Age`, to be called `min_age`, for each gender:**
```{r, message=FALSE}
min_age_by_sex <- q_data %>%
  group_by(Sex) %>%
  summarise(min_age = min(Age))

min_age_by_sex
```

**Extract the minimum `Age` for males, using the `filter()` method followed by 
`select()`:**
```{r}
min_age_male <- min_age_by_sex %>%
  filter(Sex == 'Male') %>%
  select(min_age) %>%
  as.numeric()

min_age_male
```

### First Way of Handling the Anomalous Point

**Omit the data corresponding to the person with this anomalous age value by 
using `filter()` to include only rows where `Age >= 17` _(as a sensible lower 
bound)_, saving the result in `q_data_omitted`:
```{r}
q_data_omitted <- q_data %>%
  filter(Age >= 17)

q_data_omitted$Age
```

**Now produce a boxplot of `Age`, stratified by `Sex`, using the data in 
`q_data_omitted`, from which the data corresponding to the person with the 
anomalous `Age` value has been omitted:**
```{r}
ggplot(q_data_omitted, aes(x = Sex, y = Age, fill = Sex)) +
  theme_light() + geom_boxplot(varwidth = 1) +
  labs(x = "Sex", y = "Age (years)",
       title = "Distribution of Age by Sex") +
  theme(legend.position = 'none')
```

### Second Way of Handling the Anomalous Point

**Calculate the median `Age` of all males who didn't provide an unrealistic age, 
producing a numerical result using `as.numeric()`:**
```{r, message=FALSE}
median_age_male <- q_data_omitted %>%
  group_by(Sex) %>%
  summarise(median_age = median(Age)) %>%
  filter(Sex == 'Male') %>%
  select(median_age) %>%
  as.numeric()

median_age_male
```

**Replace the unrealistic values for `Age` (`min_age_male`) with 
`median_age_male`:**
```{r}
q_data_corrected <- q_data %>%
  mutate(Age_corrected = ifelse(Age == min_age_male, median_age_male, Age))

# Check that the minimum age for males is no longer an unrealistic value:
q_data_corrected %>%
  group_by(Sex) %>%
  summarise(min_age = min(Age_corrected))
```

**Produce a boxplot of `Age`, stratified by `Sex`, using the data in 
`q_data_corrected`:**
```{r}
ggplot(q_data_corrected, aes(x = Sex, y = Age_corrected, fill = Sex)) +
  theme_light() + geom_boxplot(varwidth = T) +
  labs(x = "Sex", y = "Age (years)",
       title = "Distribution of Age by Sex",
       subtitle = "Anomalous values for age have been replaced with the median") +
  theme(legend.position = 'none')
```

The interquartile range is a measure of spread which is more robust to outliers 
than the standard deviation, and can be obtained in R using `IQR()`. **Compare 
the means, medians, standard deviations, and interquartile ranges of `Age` across 
`Sex`:**
```{r}
q_data_corrected %>%
  group_by(Sex) %>%
  summarise(mean_age = mean(Age_corrected), median_age = median(Age_corrected),
            sd_age = sd(Age_corrected), iqr_age = IQR(Age_corrected))
```
