---
title: "Session 1 - Introduction"
author: "Alex Mounsey"
date: "28/09/20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
```

## Exercise 1: Basic Data Structure, Summaries, and Plots

Consider the following data, collected by a farm management company, on the number of cattle and the number of sheep in nine farms in Devon:

$$
Cattle = \left(\begin{matrix}
  348 \\ 407 \\ 1064 \\ 750 \\ 593 \\ 1867 \\ 471 \\ 935 \\ 1443
\end{matrix}\right) \qquad
Sheep = \left(\begin{matrix}
  110 \\ 179 \\ 303 \\ 173 \\ 182 \\ 458 \\ 151 \\ 140 \\ 222
\end{matrix}\right)
$$

**Create two R objects for these data vectors, naming them `cattle` and `sheep`:**
```{r}
cattle <- c(348, 407, 1064, 750, 593, 1867, 471, 935, 1443)
sheep <- c(110, 179, 303, 173, 182, 458, 151, 140, 222)
```

**Save all observations, _except the 6th and 9th_, into two new vectors: 
`cattle_new` and `sheep_new`**
```{r}
cattle_new <- cattle[-c(6, 9)]
sheep_new <- sheep[-c(6, 9)]
```

**Create a data frame for `cattle_new` and `sheep_new`:**
```{r}
df <- data.frame(cattle_new, sheep_new)
```
```{r, echo=FALSE}
df
```

**Create a scatterplot with the values in `cattle_new` on the horizontal axis 
and the values in `sheep_new` on the vertical axis using appropriate labels 
and title, colouring the dots in blue:**
```{r}
ggplot(df, aes(x = cattle_new, y = sheep_new)) +
  theme_light() + geom_point(col = 'blue') +
  labs(x = "Number of Cattle", y = "Number of Sheep",
       title = "Livestock on Devonshire Farms")
```

**Calculate the sample mean and sample median of the values in both `cattle_new` 
and `sheep_new`:**
```{r}
df %>%
  summarise(mean_cattle_new = mean(cattle_new),
            mean_sheep_new = mean(sheep_new),
            median_cattle_new = mean(cattle_new),
            median_sheep_new = mean(sheep_new))
```

**Calculate the sum of the values in `cattle_new`, divided by the number of 
values in `cattle_new`:**
```{r}
sum_cattle_new <- sum(cattle_new)
len_cattle_new <- length(cattle_new)

sum_cattle_new / len_cattle_new
```

## Exercise 2: Factors

Ten students are asked to respond to a question: **"Overall, I was satisfied 
with my experience of this module"** on a five point scale: _"Strongly Disagree"_, 
_"Disagree"_, _"Neutral"_, _"Agree"_, and _"Strongly Agree"_. The results were 
recorded numerically using the following encoding:

| Numerical Encoding | Corresponding Response |
|:------:|:------:|
| 1 | Strongly Disagree |
| 2 | Disagree |
| 3 | Neutral |
| 4 | Agree |
| 5 | Strongly Agree |

Here are the results from the ten students:
```{r}
results <- c(5, 5, 2, 4, 3, 5, 5, 1, 2, 5)
```

**Convert these results into a factor with the labels: _"Strongly Disagree"_, 
_"Disagree"_, _"Neutral"_, _"Agree"_, and _"Strongly Agree"_:**
```{r}
results_factor <- factor(results, levels = c(1, 2, 3, 4, 5),
                         labels = c('Strongly Disagree', 'Disagree', 'Neutral', 'Agree',
                                    'Strongly Agree'))
```

**Tabulate these results:**
```{r}
results_table <- table(results_factor)
```
```{r, echo=FALSE}
results_table
```

**Now, create a factor with the labels in the reverse order, and tabulate it:**
```{r}
reversed_results_factor <- factor(results, levels = c(5, 4, 3, 2, 1),
                                  labels = c('Strongly Agree', 'Agree', 'Neutral',
                                             'Disagree', 'Strongly Disagree'))

reversed_results_table <- table(reversed_results_factor)
```
```{r, echo=FALSE}
reversed_results_table
```

## Exercise 3: Somewhat Harder

A small company has recorded the following data on production volume _(in 1,000s 
of units)_ and production costs _(in £1,000s)_ for the past ten months:
```{r}
volume <- c(2, 4, 6, 6, 10, 8, 5, 7, 11, 12)
costs <- c(7, 11, 12, 19, 22, 20, 16, 13, 24, 20)
```

**Create a data frame containing the variables `volume` and `costs`, before 
plotting this data. Comment on your plot:**
```{r}
# Create a data frame, using the provided vectors
df <- data.frame(volume, costs)

# Plot production volume vs. production costs
ggplot(df, aes(x = volume, y = costs)) +
  theme_light() + geom_point() +
  labs(x = "Production Volume (1000s of units)",
       y = "Production Costs (£1,000s)",
       title = "Production Volume vs. Production Costs")
```

### Comments:
There appears to be a positive linear relationship between the number of units 
produced and the production costs. This is to be expected: producing more units 
is going to cost more money (labor, materials, etc.).

**Create a time-series plot of production volume, and comment on this plot:**
```{r}
# Create a vector of months and add it to the data frame
month <- c(1:10)
df <- cbind(df, month)

# Plot production volume vs. time
ggplot(df, aes(x = month, y = volume)) +
  theme_light() + geom_line() +
  labs(x = "Time (months)", y = "Production Volume (1000s of units)",
       title = "Production Volume over Time")
```

### Comments:
The production volume steadily increases over time, this is likely due to the 
expansion of the business as they gain more clients. There appears to be dip in 
production volume between months `5` and `8`.

**Create a time-series plot of production costs, and commend on this plot:**
```{r}
ggplot(df, aes(x = month, y = costs)) +
  theme_light() + geom_line() +
  labs(x = "Time (months)", y = "Production Costs (£1,000s)",
       title = "Production Costs over Time")
```

### Comments:
As you might expect, the production costs increase over time in line with the 
production volume.

**Create a plot showing production volume and production costs, with the month 
represented as text-based data points:**
```{r}
ggplot(df, aes(x = volume, y = costs, label = row.names(df))) +
  theme_light() + geom_text() +
  labs(x = "Production Volume (1000s of units)",
       y = "Production Costs (£1,000s)",
       title = "Production Volume vs. Production Costs over Time",
       subtitle = "Months are represented as the data points' text")
```
