---
title: "Session 4 - ggplot2"
author: "Alex Mounsey"
date: "07/10/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(ggplot2)
library(readxl)
library(scales)
library(tidyr)
```


## Exercise 1

Consider again the data collected by a farm management company from nine farms 
in Devon about the number of cattle and sheep. It turns out that other 
information is available, including the farm location and whether the opportunity 
to participate in a working farm holiday is offered:

$$
Cattle = \left(\begin{matrix}
  348 \\ 407 \\ 1064 \\ 750 \\ 593 \\ 1867 \\ 471 \\ 935 \\ 1443
\end{matrix}\right) \qquad
Sheep = \left(\begin{matrix}
  110 \\ 179 \\ 303 \\ 173 \\ 182 \\ 458 \\ 151 \\ 140 \\ 222
\end{matrix}\right) \qquad
Location = \left(\begin{matrix}
  North \\ South \\ South \\ North \\ South \\ North \\ North \\ North \\ South
\end{matrix}\right) \qquad
Holidays = \left(\begin{matrix}
  No \\ No \\ Yes \\ Yes \\ No \\ Yes \\ No \\ No \\ Yes
\end{matrix}\right)
$$

```{r}
cattle <- c(348, 407, 1064, 750, 593, 1867, 471, 935, 1443)
sheep <- c(110, 179, 303, 173, 182, 458, 151, 140, 222)
location <- c('North', 'South', 'South', 'North', 'South', 'North', 'North',
              'North', 'South')
holidays <- c('No', 'No', 'Yes', 'Yes', 'No', 'Yes', 'No', 'No', 'Yes')
```

**Define categorical variables `location` and `holidays` as factors and create 
a dataframe including these newly defined variables, as well as the `cattle` 
and `sheep` variables that you've worked with before:**

```{r}
location_f <- factor(location, levels = c('North', 'East', 'South', 'West'))
holidays_f <- factor(holidays, levels = c('Yes', 'No'))

farms_df <- data.frame(cattle, sheep, location_f, holidays_f) %>%
  rename(location = location_f, holidays = holidays_f)

head(farms_df)
```

**Reproduce the plots shown in the tutorial file:**

```{r, message=FALSE}
ggplot(farms_df, aes(x = cattle, y = sheep)) +
  theme_light() + geom_point() + geom_smooth(span = 1) +
  labs(x = "Number of Cattle", y = "Number of Sheep",
       title = "Livestock on Devonshire Farms")
```

```{r, message=FALSE}
ggplot(farms_df, aes(x = cattle, y = sheep, colour = holidays)) +
  theme_light() + geom_point() + geom_smooth(method = 'lm', se = F) +
  facet_grid(holidays ~ location) +
  labs(x = "Number of Cattle", y = "Number of Sheep",
       colour = "Working Farm Holidays",
       title = "Livestock on Devonshire Farms",
       subtitle = "Faceted by location and working farm holidays")
```

## Exercise 2

**Read the questionnaire data into a dataframe, `df`, in the usual way:**

```{r, message=FALSE}
q_data <- read_csv('../data/MATH513_Questionnaire_Data.csv')
head(q_data)
```

**Using the `dplyr` package, select and display the columns _(variables)_ 
`Travel_time` and `Distance`:**

```{r}
q_data %>%
  select(Travel_time, Distance) %>%
  head()
```

**Your goal is to understand how `Travel_time` is dependent on `Distance`. Use 
`ggplot2` to plot `Travel_time` against `Distance`:**

```{r, message=FALSE}
ggplot(q_data, aes(x = Distance, y = Travel_time)) +
  theme_light() +
  geom_point() + geom_smooth() + geom_smooth(method = 'lm', se = F, colour = 'red') +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Distance (miles)", y = "Travel Time (minutes)",
       title = "Distance vs. Travel Time")
```

### Logarithmic Axis Scales

The logarithmic scale is used in situations where data points range between 
extremely large values and extremely small values, such that the detail of the 
smaller values is not lost.

### Regression Line

The general mathematical equation of a straight line is:
$$
y = \alpha + \beta x
$$
where: $\alpha$ is known as the intercept and $\beta$ is known as the slope.

When we're handling data, we can't expect the data points to lie perfectly on 
the line, so we allow an error. The simple linear regression model therefore 
takes the form:

$$
y = \alpha + \beta x + error
$$

## Exercise 3

The file `companies.xlsx` contains information collected from 100 companies 
belonging to different sectors of the economy. The following variables have been 
recorded:

* `company`: The company's ID
* `net_income_2015`: The company's net income in 2015
* `net_income_2014`: The company's net income in 2014
* `oper_result`: The company's operational result in 2015
* `lab_cost`: The company's labor expenditure in 2015
* `n_empl`: The number of employees in the company
* `sector`: The economic sector the company operates within

```{r}
c_data <- read_excel('../data/companies.xlsx')
head(c_data)
```

**Create a new variable, `net_income_diff`, as the difference between the 2015 
net income nad the 2014 net income:**

```{r}
c_data <- c_data %>%
  mutate(net_income_diff = net_income_2015 - net_income_2014)

head(c_data$net_income_diff)
```

**Produce a scatterplot of the net income difference _(on the vertical axis)_ 
against the number of employees _(on the horizontal axis)_. Add a smooth curve 
to your scatter plot.

```{r, message=FALSE}
ggplot(c_data, aes(x = n_empl, y = net_income_diff)) +
  theme_light() + geom_point() + geom_smooth(span = 1) +
  scale_y_continuous(label = comma) +
  labs(x = "Number of Employees", y = "Net Income Difference (2014-2015)",
       title = "Number of Employees vs. Net Income Difference from 2014 to 2015")
```

**Using only the companies with a positive number of employees and a positive 
net income difference, modify the above plot to use a logarithmic scale on 
both axis:**

```{r, message=FALSE}
filter(c_data, n_empl > 0 & net_income_diff > 0) %>%
  ggplot(aes(x = n_empl, y = net_income_diff)) +
    theme_light() + geom_point() + geom_smooth(span = 1) +
    scale_x_log10(label = comma) + scale_y_log10(label = comma) +
    labs(x = "Number of Employees", y = "Net Income Difference (2014-2015)",
         title = "Number of Employees vs. Net Income Difference from 2014 to 2015")
```

**Using all of the companies, create a factor, `sector_new`, based on the 
variable `sector`, such that the levels of the factor are labelled `catering`, 
_(corresponding to 1)_, `hotels` _(corresponding to 2)_, `distribution` 
_(corresponding to 3)_, and `communications` _(corresponding to 4)_:**

```{r}
c_data <- c_data %>%
  mutate(sector_new = factor(sector, levels = 1:4,
                             labels = c('Catering', 'Hotels', 'Distribution',
                                        'Communications')))

head(c_data$sector_new)
```

**Using all of the companies, produce a boxplot of `n_empl` against `sector_new`:**

```{r, warning=FALSE}
ggplot(c_data, aes(x = sector_new, y = n_empl, colour = sector_new)) +
  theme_light() + geom_boxplot() +
  scale_y_log10(label = comma) +
  labs(x = "Sector", y = "Number of Employees",
       title = "Economic Sector vs. Number of Employees") +
  theme(legend.position = 'none')
```

**Using all of the companies, convert the continuous variable, `oper_result`, 
_(which has units of GBP)_ into the continous variable `oper_result_millions` 
_(with units of millions of GBP)_:**

```{r}
c_data <- c_data %>%
  mutate(oper_result_millions = oper_result / 1000000)

head(c_data$oper_result_millions)
```

**Using all of the companies, convert the continuous variable `oper_result_millions` 
into the factor `oper_result_new`, with breaks at -5, 0, 0.25, and 3, in such a 
way that the corresponding intervals are |-5, 0), |0, 0.25) and |0.25, 3):**

```{r}
c_data <- c_data %>%
  mutate(oper_result_new = cut(x = oper_result_millions, breaks = c(-5, 0, 0.25, 3)))

table(c_data$oper_result_new)
```

**Using all of the companies, produce histograms of `n_empl` split by 
`oper_result_new`, using a logarithmic scale for the number of employees:**

```{r, message=FALSE, warning=FALSE}
ggplot(c_data, aes(x = n_empl, fill = oper_result_new)) +
  theme_light() + geom_histogram() +
  scale_x_log10(label = comma) +
  facet_grid(oper_result_new ~ .) +
  labs(x = "Number of Employees", y = "Count", fill = "Operational Result (millions £)",
       title = "Number of Employees by Operational Result")
```

## Exercise 4: Displaying Student Assessment Data

The file `Module_Marks_Invented_Example.csv` contains some student assessment 
data. It would not be ethically correct to release students' marks, so these 
marks are entirely fabricated, although they do share some of the properties 
of real marks.

Students take three modules: M1, M2, and M3. The coursework/examination/overall 
marks are indicated by the suffix of the column title _(.C/.E/.F respectively)_.

```{r, message=FALSE}
m_data <- read_csv('../data/Module_Marks_Invented_Example.csv')
head(m_data)
```

Note that as not all students take all modules, there are a lot of missing values.

**Use the `gather()` function from the `tidyr` package to place all of the marks 
into one column names `marks`, with another column named `source` indicating 
the source of the marks _(i.e. M1.C, M1.E, M1.F, M2.C, M2.E, M2.F, M3.C, M3.E, 
M3.F)_:**

```{r}
m_long <- gather(m_data, 'source', 'marks')
head(m_long)
```

**Use the `separate()` function from the `tidyr` package to separate the `source` 
column of `m_long` into two columns named `module` _(containing the module code 
M1, M2, or M3)_ and `component` _(containing the assessment component C, E, or F)_. 
Do not change the `marks` column:**

```{r}
m_long_2 <- m_long %>%
  separate(col = source, into = c('module', 'component'), sep = '\\.')
head(m_long_2)
```

**Use the `mutate()` function from `dplyr` to define `component` as a factor, 
named `component_f`, with levels: "C", "E", and "F" and labels: "Coursework", 
"Examination", and "Overall":**

```{r}
m_long_3 <- m_long_2 %>%
  mutate(component_f = factor(component, levels = c('C', 'E', 'F'),
                              labels = c('Coursework', 'Examination', 'Overall')))
head(m_long_3$component_f)
```

**Finally, produce a boxplot of this data using `ggplot2`:**

```{r, warning=FALSE}
ggplot(m_long_3, aes(x = component_f, y = marks, fill = component_f)) +
  theme_light() + geom_boxplot() + facet_grid(. ~ module) +
  scale_fill_manual(values = c('orange', 'yellow', 'magenta')) +
  scale_y_continuous(breaks = c(30, 40, 50, 60, 70, 100),
                     minor_breaks = c(30, 40, 50, 60, 70, 80, 90, 100),
                     limits = c(30, 100)) +
  labs(x = "Component", y = "Mark (%)") +
  theme(axis.text.x = element_text(size = 14, angle = 90, vjust = 0.5, hjust = 1),
        legend.position = 'none')
```
